\pkg{SWMPr}: An R Package for Retrieving, Organizing, and Analyzing Environmental Data for Estuaries

### 2015-09-03 Rjournal/beck.tex
submitted to R Journal

\documentclass[a4paper]{report}\usepackage[]{graphicx}\usepackage{xcolor}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%

\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{RJournal}
\usepackage{amsmath,amssymb,array}
\usepackage{booktabs}

%% load any required packages

% bold/italic text
\newcommand{\Bigtxt}[1]{\textbf{\textit{#1}}}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}

%% do not edit, for illustration only
\sectionhead{Contributed research article}
\volume{XX}
\volnumber{YY}
\year{20ZZ}
\month{AAAA}

\begin{article}

%knitr options

@@

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
# retrieve metadata for all sites
site_codes()

# retrieve metadata for a single site
site_codes_ind('apa')
\end{verbatim}
\end{kframe}
\end{knitrout}

Retrieval functions to import data directly into R from the CDMO include \code{all\_params}, \code{all\_params\_dtrng}, and \code{single\_param}. Due to rate limitations on the CDMO server, the retrieval functions return a limited number of records with each request.  However, the \pkg{SWMPr} functions use the native CDMO web services iteratively (i.e., within a loop) to obtain all requested records.  Download time can be excessive for longer time series.     
\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
# all parameters for a station, most recent
all_params('hudscwq')

@@

# single parameter for a station, most recent
single_param('hudscwq', param = 'do_mgl')
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
# import all paramaters for the station
# three most recent records
exdat <- all_params('apadbwq', Max = 3, trace = F)
exdat
##         datetimestamp temp f_temp spcond f_spcond sal f_sal do_pct
## 1 2015-08-18 07:00:00   28      0     34        0  21     0     77
## 2 2015-08-18 07:15:00   28      0     34        0  21     0     75
## 3 2015-08-18 07:30:00   28      0     33        0  21     0     82
##   f_do_pct do_mgl f_do_mgl depth f_depth ph f_ph turb f_turb chlfluor
## 1        0      5        0     2       0  8    0   13      0       NA
## 2        0      5        0     2       0  8    0    9      0       NA
## 3        0      6        0     2       0  8    0   11      0       NA
##   f_chlfluor level f_level cdepth clevel f_cdepth f_clevel
## 1         -2    NA      -1     NA     NA       -1         
## 2         -2    NA      -1     NA     NA       -1         
## 3         -2    NA      -1     NA     NA       -1
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
# import sample data from package
data(apadbwq)
dat <- apadbwq

@@

# view a single attribute of dat
attr(dat, 'station')
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
# view available methods for swmpr class
methods(class = 'swmpr')
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
# qaqc screen for a swmpr object, retain only '0'
qaqc(dat)

@@

# retain only '0' and '-1' flags
qaqc(dat, qaqc_keep = c(0, -1))
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
# view the number of observations in each QAQC flag
qaqcchk(dat)
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
# import data
data(apaebmet)
dat <- apaebmet

@@

# subset records within a date range, select two parameters
subset(dat, subset = c('2012-07-01 6:00', '2012-08-01 18:15'),
  select = c('atemp', 'totsorad'))
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
# import, qaqc removal
data(apadbwq)
dat <- qaqc(apadbwq)

# convert time series to two hour invervals
# tolerance of +/- 30 minutes for matching existing data
setstep(dat, timestep = 120, differ = 30)
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
# get nut, wq, and met data as separate objects
data(apacpnut)
data(apacpwq)
data(apaebmet)
swmp1 <- apacpnut
swmp2 <- apacpwq
swmp3 <- apaebmet

@@

# combine nut, wq, and met data by nut time series, two hour time step
comb(swmp1, swmp2, swmp3, timestep = 120, method = 'apacpnut')
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
# get data
data(apadbwq)
dat <- apadbwq

@@

# daily decomposition of DO and plot
dc_dat <- decomp(dat, param = 'do_mgl', frequency = 'daily')
plot(dc_dat)
\end{verbatim}
\end{kframe}\begin{figure}[!h]

{\centering \includegraphics[width=\textwidth]{figure/decomp_ex1-1}

@@

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
# get data
data(apacpnut)
dat <- apacpnut
dat <- qaqc(dat, qaqc_keep = NULL)

# decomposition of chl
decomp_cj(dat, param = 'chla_n')
\end{verbatim}
\end{kframe}\begin{figure}[!h]

{\centering \includegraphics[width=\textwidth]{figure/decomp_ex2-1}

@@

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
## import water quality and weather data
data(apadbwq)
data(apaebmet)

@@

## estimate metabolism
res <- ecometab(dat, trace = FALSE)
plot_metab(res)
\end{verbatim}
\end{kframe}\begin{figure}[!h]

{\centering \includegraphics[width=\textwidth]{figure/metab_ex-1}

@@

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
## import data
data(apacpnut)
dat <- qaqc(apacpnut)

## plot
plot_summary(dat, param = 'chla_n', years = c(2007, 2013))
\end{verbatim}
\end{kframe}\begin{figure}[!h]

{\centering \includegraphics[width=\textwidth]{figure/summary_ex-1}

@@

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
## import data
data(apacpwq)
dat <- qaqc(apacpwq)

## plot
overplot(dat, select = c('depth', 'do_mgl', 'ph', 'turb'),
  subset = c('2013-01-01 0:0', '2013-02-01 0:0'), lwd = 2)
\end{verbatim}
\end{kframe}\begin{figure}[!h]

{\centering \includegraphics[width=\textwidth]{figure/overplot-1}

@@

\begin{knitrout}\small
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
# plot the stations at Jacques Cousteau reserve
map_reserve('jac')
\end{verbatim}
\end{kframe}\begin{figure}[!h]

{\centering \includegraphics[width=\textwidth]{figure/map_ex-1}

@@

\begin{figure}
\begin{center}
\includegraphics[width = \textwidth]{figure/swmp_comp.pdf}
\caption{Online application for comparing trends in SWMP data parameters using an interactive map.  Link: \href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}}
\label{fig:swmp_comp}
\end{center}
\end{figure}

@@

\end{article}

\end{document}

>>>>

@@

\begin{example}
# retrieve metadata for all sites
site_codes()

# retrieve metadata for a single site
site_codes_ind('apa')
\end{example}

Retrieval functions to import data directly into R from the CDMO include \code{all\_params}, \code{all\_params\_dtrng}, and \code{single\_param}. Due to rate limitations on the CDMO server, the retrieval functions return a limited number of records with each request.  However, the \pkg{SWMPr} functions use the native CDMO web services iteratively (i.e., within a loop) to obtain all requested records.  Download time can be excessive for longer time series.     
\begin{example}
# all parameters for a station, most recent
all_params('hudscwq')

@@

# single parameter for a station, most recent
single_param('hudscwq', param = 'do_mgl')
\end{example}

@@

\begin{example}
# import all paramaters for the station
# three most recent records
exdat <- all_params('apadbwq', Max = 3, trace = F)
exdat
##         datetimestamp temp f_temp spcond f_spcond sal f_sal do_pct
## 1 2015-08-18 07:00:00   28      0     34        0  21     0     77
## 2 2015-08-18 07:15:00   28      0     34        0  21     0     75
## 3 2015-08-18 07:30:00   28      0     33        0  21     0     82
##   f_do_pct do_mgl f_do_mgl depth f_depth ph f_ph turb f_turb chlfluor
## 1        0      5        0     2       0  8    0   13      0       NA
## 2        0      5        0     2       0  8    0    9      0       NA
## 3        0      6        0     2       0  8    0   11      0       NA
##   f_chlfluor level f_level cdepth clevel f_cdepth f_clevel
## 1         -2    NA      -1     NA     NA       -1         
## 2         -2    NA      -1     NA     NA       -1         
## 3         -2    NA      -1     NA     NA       -1
\end{example}

@@

\begin{example}
# import sample data from package
data(apadbwq)
dat <- apadbwq

@@

# view a single attribute of dat
attr(dat, 'station')
\end{example}

@@

\begin{example}
# view available methods for swmpr class
methods(class = 'swmpr')
\end{example}

@@

\begin{example}
# qaqc screen for a swmpr object, retain only '0'
qaqc(dat)

@@

# retain only '0' and '-1' flags
qaqc(dat, qaqc_keep = c(0, -1))
\end{example}

@@

\begin{example}
# view the number of observations in each QAQC flag
qaqcchk(dat)
\end{example}

@@

\begin{example}
# import data
data(apaebmet)
dat <- apaebmet

@@

# subset records within a date range, select two parameters
subset(dat, subset = c('2012-07-01 6:00', '2012-08-01 18:15'),
  select = c('atemp', 'totsorad'))
\end{example}

@@

\begin{example}
# import, qaqc removal
data(apadbwq)
dat <- qaqc(apadbwq)

# convert time series to two hour invervals
# tolerance of +/- 30 minutes for matching existing data
setstep(dat, timestep = 120, differ = 30)
\end{example}

@@

\begin{example}
# get nut, wq, and met data as separate objects
data(apacpnut)
data(apacpwq)
data(apaebmet)
swmp1 <- apacpnut
swmp2 <- apacpwq
swmp3 <- apaebmet

@@

# combine nut, wq, and met data by nut time series, two hour time step
comb(swmp1, swmp2, swmp3, timestep = 120, method = 'apacpnut')
\end{example}

@@

\begin{example}
# get data
data(apadbwq)
dat <- apadbwq

@@

# daily decomposition of DO and plot
dc_dat <- decomp(dat, param = 'do_mgl', frequency = 'daily')
plot(dc_dat)
\end{example}
\begin{figure}[!h]

{\centering \includegraphics[width=\textwidth]{decomp_ex1-1}

@@

\begin{example}
# get data
data(apacpnut)
dat <- apacpnut
dat <- qaqc(dat, qaqc_keep = NULL)

# decomposition of chl
decomp_cj(dat, param = 'chla_n')
\end{example}
\begin{figure}[!h]

{\centering \includegraphics[width=\textwidth]{decomp_ex2-1}

@@

\begin{example}
## import water quality and weather data
data(apadbwq)
data(apaebmet)

@@

## estimate metabolism
res <- ecometab(dat, trace = FALSE)
plot_metab(res)
\end{example}
\begin{figure}[!h]

{\centering \includegraphics[width=\textwidth]{metab_ex-1}

@@

\begin{example}
## import data
data(apacpnut)
dat <- qaqc(apacpnut)

## plot
plot_summary(dat, param = 'chla_n', years = c(2007, 2013))
\end{example}
\begin{figure}[!h]

{\centering \includegraphics[width=\textwidth]{summary_ex-1}

@@

\begin{example}
## import data
data(apacpwq)
dat <- qaqc(apacpwq)

## plot
overplot(dat, select = c('depth', 'do_mgl', 'ph', 'turb'),
  subset = c('2013-01-01 0:0', '2013-02-01 0:0'), lwd = 2)
\end{example}
\begin{figure}[!h]

{\centering \includegraphics[width=\textwidth]{overplot-1}

@@

\begin{example}
# plot the stations at Jacques Cousteau reserve
map_reserve('jac')
\end{example}
\begin{figure}[!h]

{\centering \includegraphics[width=\textwidth]{map_ex-1}

@@

\begin{figure}
\begin{center}
\includegraphics[width = \textwidth]{swmp_comp.pdf}
\caption{Online application for comparing trends in SWMP data parameters using an interactive map.  Link: \href{https://beckmw.shinyapps.io/swmp_comp}{https://beckmw.shinyapps.io/swmp\_comp}}
\label{fig:swmp_comp}
\end{center}
\end{figure}

@@

### 2015-09-03 cover_letter.tex
submitted to R Journal

\emph{To:}\newline
Dr. Damian Pattinson\newline
Editorial Director\newline
PLOS ONE

\emph{From:}\newline
Dr. Marcus W. Beck\newline
ORISE Research Participation Program\newline
US Environmental Protection Agency\newline
beck.marcus@epa.gov, 850-934-2480\vspace{0.1in}

Enclosed please find my manuscript, entitled `SWMPr: An R Package for Retrieving, Organizing, and Analyzing Environmental Data for Estuaries', to be considered as a research article in PLOS ONE. A continuing challenge in the environmental sciences is the effective use of continuous monitoring data.  The increasing quantity of data from automated sensor networks suggests that more appropriate methods to synthesize and interpret this information are needed. This article describes the SWMPr (`swamper') software package that was developed to address research needs of the US National Estuarine Research Reserve System (NERRS) using the open-source statistical programming language R.  The System-Wide Monitoring Program (SWMP) was implemented by NERRS in 1995 to collect continuous water quality data at over 140 fixed stations at each of the estuaries in the reserve system.  The SWMPr package address many common issues working with large datasets from automated sensor networks, such as data pre-processing, combining data from different sources, and exploratory analyses to identify parameters of interest. Additionally, a cross-reserve comparison of water quality trends and ecosystem metabolism estimates is provided to illustrate the utility of the package.

The manuscript has not been previously submitted to PLOS ONE and I have had no prior interactions with PLOS. Please note that Dr. Jeffrey Hollister (US EPA) has already provided a review of an earlier draft. I greatly appreciate the opportunity to publish my work in PLOS ONE.  Below is a suggested list of appropriate editors to handle my manuscript.

Zaid Abdo,
Institution and Department: Agricultural Research Service,
UNITED STATES

Ruth H. Carmichael,
Dauphin Island Sea Lab; University of South Alabama,
UNITED STATES

Juan Carlos Molinero,
GEOMAR: Helmholtz Center for Ocean Research,
GERMANY

Tomoya Iwata,
University of Yamanashi,
JAPAN

>>>>

\emph{To:}\newline
Bettina Gr\"un\newline
Editor in Chief\newline
The R Journal

\emph{From:}\newline
Marcus W. Beck\newline
ORISE Research Participation Program\newline
US Environmental Protection Agency\newline
beck.marcus@epa.gov, 850-934-2480\vspace{0.1in}

Enclosed please find my manuscript, entitled `SWMPr: An R Package for Retrieving, Organizing, and Analyzing Environmental Data for Estuaries', to be considered as a research article in the R Journal.

A continuing challenge in the environmental sciences is the effective use of continuous monitoring data.  The increasing quantity of data from automated sensor networks suggests that more appropriate methods to synthesize and interpret this information are needed. This article describes the SWMPr (`swamper') software package for R that was developed in part to address research needs of the US National Estuarine Research Reserve System (NERRS).  The System-Wide Monitoring Program (SWMP) was implemented by NERRS in 1995 to collect continuous water quality data at over 140 fixed stations at each of the estuaries in the reserve system.  The SWMPr package addresses many common issues working with large datasets from automated sensor networks, such as data pre-processing, combining data from different sources, and exploratory analyses to identify parameters of interest. As such, the package has value for the NERRS network as well as more generic applications to environmental time series data.  Several functions provided in the package are novel contributions to CRAN.

I greatly appreciate the opportunity to publish my work in the R Journal.  Please feel free to contact me directly should additional information be needed.
